---
title: "Class 10: Halloween Mini-Project"
author: "Heidi Nam"
format: html
editor: visual
---

## Importing candy data

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

Q1: How many different candy types are in this dataset?

```{r}
ncol(candy)
```

-   There are 12 different types of candies.

Q2: How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

-   There are 38 fruity candy types in the dataset.

## 2. What is your favorite candy?

Looking at `winpercent` through my favorite choice of candy:

Q3: what is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}
candy["Nerds", ]$winpercent
```

-   The `winpercent` value of Nerds is 55.35405.

Q4: What is the `winpercent` value for Kit Kat?

```{r}
candy["Kit Kat", ]$winpercent
```

-   The `winpercent` value of Kit Kat is 76.7686.

Q5: What is the `winpercent` value for Tootsie Roll Snack Bars?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

-   The `winpercent` value of Tootsie Roll Snack Bars is 49.6535.

Understanding `skim()` function:

```{r}
library("skimr")
skim(candy)
```

Q6: Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

-   The `winpercent` value is on a different scale as its mean does not rely from a value from 0 to 1 but from 0 to 100.

Q7: What do you think a zero and one represent for the `candy$chocolate` column?

-   The zero most likely represents that the candy does not contain chocolate while the one represents that the candy does contain chocolate.

Starting exploratory analysis with histogram:

Q8: Plot a histogram of `winpercent` values

```{r}
hist(candy$winpercent)
```

Q9: Is the distribution of `winpercent` values symmetrical?

-   no, it is slightly positive skewed.

Q10: Is the center of the distribution above or below 50%?

-   The center of distribution is below 50%.

Q11: On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolatewin <- candy$winpercent[as.logical(candy$chocolate)]
fruitywin <- candy$winpercent[as.logical(candy$fruity)]
mean(chocolatewin)
mean(fruitywin)
```

-   On average, chocolate candy is higher ranked than fruity candy.

Q12: Is this difference statistically significant?

```{r}
t.test(chocolatewin, y = fruitywin)
```

-   as the p-value is smaller than 0.05, the difference is statistically significant.

## 3. Overall candy rankings

Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

-   The five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble and Jawbusters.

Q14: What are the top all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```

-   The five most liked candies are Reese's Peanut Butter, Reese's Miniatures, Twix, Kit Kat and Snickers.

Q15: Make a first `barplot` of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(data = candy) +
  aes(winpercent, rownames(candy)) + 
  geom_bar(stat='identity')
```

Q16. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
ggplot(data = candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) + 
  geom_bar(stat='identity')
```

Adding color to the set:

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(data = candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) + 
  geom_col(fill=my_cols)
```

Q17: What is the worst ranked chocolate candy?

-   Charleston Chew

Q18: What is the best ranked fruity candy?

-   Starburst

## 4. Taking a look at `pricepercent`

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

Q19: Which candy type is the highest ranked in terms of `winpercent` for the least money?

-   Reese's miniatures

Q20: What are the top 5 most expensive candy types in the data set and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

-   The least popular most expensive candy would be Nik L Nip.

## 5 Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

Q22: Examining this plot what two variables are anti-correlated?

-   fruity and chocolate are anti-correlated.

Q23: Similarly, what two variables are most positively correlated?

-   chocolate and winpercent are positively correlated, and so is chocolate and bar.

## 6. Principal component analysis

applying PCA using the `prcomp()` function:

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

plotting PC1 and PC2 of PCA

```{r}
plot(pca$x[,])
```

changing plotting character:

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

making it better with ggplot2:

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

labeling the plot with ggrepel:

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```

using plotly to generate an interactive plot:

```{r}
# install.packages("plotly")
library(plotly)
```

```{r}
ggplotly(p)
```

taking a look at PCA:

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Q24: What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

-   Fruity, hard and pluribus variables are picked up strongly by PC1. This makes sense as many fruity candy are hard and come as a batch of a multiple of them when bought.
